---
title: "Mental Health and Economic Inequality in High-Income Countries"
author: "Alona Sychevska, 2769408"
date: "June 5, 2025"
output: html_document
---

```{r warning=FALSE, message = F}
library(psych)
library(GGally)
library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(scales)
```
# Title page

Alona Sychevska

Tutorial lecturer's name: Chantal Schouwenaar

# Part 1 - Identify a Social Problem

## 1.1 Describe the Social Problem

Mental health issues such as depression and anxiety are rising in high-income countries. Countries with excellent healthcare infrastructure—like the Netherlands, Germany, or the UK—report increasing rates of psychological distress. 

This project explores whether **economic inequality** (measured by Gini coefficient, poverty rate, etc.) helps explain this rise. It builds on the growing body of research suggesting that **relative deprivation** and **social exclusion** may drive mental illness more than absolute income levels.

**Sources:**
- WHO Global Health Observatory (depression, anxiety)

- OECD (Gini index, poverty rate)

- Eurostat (health expenditure, mental health indicators)

- Wilkinson & Pickett (The Spirit Level)

---

# EDA (Exploratory Data Analysis)

# Part 2 - Data Sourcing

## 2.1 Load in the data

```{r loading data}
mental<- read_csv("/home/qgelena/code/VU/R_projects/data/Global-Burden-of-Disease-Study/Global-Burden-of-Disease-Study.csv", show_col_types = FALSE)

inequality <- read_csv("/home/qgelena/code/VU/R_projects/data/GINI/GINI.csv", show_col_types = FALSE)
```
# 2.2 Checking the structure of the dataset mental health

```{r}
dim(mental)
names(mental)
head(mental)

unique(mental$measure_name)
```

# 2.2 Checking the structure of the dataset inequality

```{r}
dim(inequality)
names(inequality)
head(inequality)

```

# 2.2 Provide a short summary of the dataset(s)
mental: Depression/anxiety prevalence per 100,000 people

inequality: Gini index, poverty rate

expenditure: Health spending per capita / % GDP

Time period: 2000–2023

Countries: Netherlands, UK, Germany, Sweden, US, Japan, etc.

# 2.3 Describe the type of variables included

1. age: age in years, непрерывный признак
2. sex (1 = male; 0 = female), качественный признак
3. cp: chest pain type, сердечно сосудистое заболевание, качественный
-- Value 1: typical angina
-- Value 2: atypical angina
-- Value 3: non-anginal pain
-- Value 4: asymptomatic
4. trestbps: resting blood pressure (in mm Hg on admission to the hospital), непрерывный
5. chol: serum cholestoral in mg/dl, сывороточный холестерин, непрерывный
6. fbs: (fasting blood sugar > 120 mg/dl) (1 = true; 0 = false), сахар в крови натощак, качественный
7. restecg: resting electrocardiographic results, качественный
-- Value 0: normal
-- Value 1: having ST-T wave abnormality (T wave inversions and/or ST elevation or depression of > 0.05 mV)
-- Value 2: showing probable or definite left ventricular hypertrophy by Estes' criteria
8. thalach: maximum heart rate achieved, непрерывный
9. exang:  exercise induced angina (1 = yes; 0 = no), качественный
10. oldpeak = ST depression induced by exercise relative to rest, углубление/опускание ST во время нагрузки, сравнительно с отдыхом
11. slope: the slope of the peak exercise ST segment, наклон ST сегмента во время пиковой нагрузки, качественный
-- Value 1: upsloping
-- Value 2: flat
-- Value 3: downsloping
12. ca: number of major vessels (0-3) colored by flourosopy, дисретный
13. thal (Thalium, a radioactive tracer injected during a stress test *): 3 = normal; 6 = fixed defect; 7 = reversable defect, качественный
14. num: diagnosis of heart disease (angiographic disease status), качественный
-- Value 0: < 50% diameter narrowing
-- Value 1-4: > 50% diameter narrowing
(in any major vessel: attributes 59 through 68 are vessels)



Mental health: WHO modeled estimates, % of population affected

Inequality: administrative data from OECD

Spending: macroeconomic statistics

# Part 3 - Quantifying

## 3.1 Data cleaning 

# Check missed data mental health dataset

```{r}
colSums(is.na(mental))
```

# Check missed data inequality dataset

```{r}
colSums(is.na(inequality))
```
## Generate necessary variables

```{r}
mental_not_death <- mental[mental$measure_name == "DALYs (Disability-Adjusted Life Years)",]
mental_not_death <- subset(mental_not_death, location_name %in% c(
  "Norway",
  "Sweden",
  "Netherlands",
  "Germany",
  "France",
  "United Kingdom",
  "United States",
  "Canada",
  "Australia",
  "Japan",
  "South Korea",
  "Italy",
  "Spain",
  "Finland",
  "Belgium")
  )
unique(mental$cause_name)

#only depression and anxiety
not_death_dep_anx <- mental_not_death[
  mental_not_death$sex_id == 3 & 
  mental_not_death$age_id == 22 &
  (mental_not_death$cause_name == "Anxiety disorders" | 
   mental_not_death$cause_name == "Depressive disorders"), 
]

mental_NL <- subset(not_death_dep_anx, location_name == "Netherlands")


mental_not_death <- mental_not_death[mental_not_death$cause_name == "All causes" & mental_not_death$sex_id == 3 & mental_not_death$age_id == 22, ]
head(mental_not_death)

mental_Sweden <- subset(mental_not_death, location_name == "Sweden")
mental_France <- subset(mental_not_death, location_name == "France")

```
## Working on inequality dataset


## Data cleaning
# filter data by country
```{r}
inequality <- subset(inequality, `Country Name` %in% c(
  "Norway",
  "Sweden",
  "Netherlands",
  "Germany",
  "France",
  "United Kingdom",
  "United States",
  "Canada",
  "Australia",
  "Japan",
  "South Korea",
  "Italy",
  "Spain",
  "Finland",
  "Belgium")
  )
```
# transform dataset in the way years are not the name of columns anymore:
```{r}
inequality_long <- inequality %>%
  pivot_longer(
    cols = matches("^\\d{4}$"),
    names_to = "year",
    values_to = "gini_index"
  ) %>%
  mutate(year = as.integer(year))
```
# Pick the latest available gini-index
```{r}
latest_gini <- inequality_long %>%
  arrange(`Country Name`, desc(year)) %>%
  group_by(`Country Name`) %>%
  filter(!is.na(gini_index)) %>%
  slice(1) %>%
  ungroup()

head(latest_gini)
```
# Create buckets
```{r}
gini_buckets <- latest_gini%>%
  mutate(gini_bucket = case_when(
    gini_index <25 ~ "Low",
    gini_index >= 25 & gini_index <35 ~ "Middle",
    gini_index >= 35 ~ "High",
    TRUE ~ NA_character_
  ))
```

## 3.3 Visualize

```{r}
#NL
ggplot(mental_NL, aes(x = year, y = val, color = cause_name)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(title = "Depression and Anxiety Cases in NL",
       x = "Year", y = "Number of Cases",
       color = "Mental Health Condition") +
  theme_minimal()

# Sweden
ggplot(mental_Sweden, aes(x = year, y = val)) +
  geom_line(color = "darkgreen", size = 1.2) +
  geom_point(color = "darkgreen", size = 3) +
  labs(title = "Mental Health Cases in Sweden Over Time",
       x = "Year", y = "Number of Cases") +
  theme_minimal()

#France
ggplot(mental_France, aes(x = factor(year), y = val)) +
  geom_col(fill = "skyblue", color = "darkblue", width = 0.7) +
  labs(title = "Mental Disorder Cases per Year in France",
       x = "Year", y = "Number of Cases") +
  scale_y_continuous(labels = label_comma()) +  # 👈 this fixes the y-axis
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```{r stacked_bar, fig.width=12, fig.height=6}
ggplot(mental, aes(x = factor(year), y = val, fill = cause_name)) +
  geom_col(position = "stack") +
  scale_y_continuous(labels = scales::comma_format()) +
  labs(
    title = "Total Mental Health Cases in France by Disorder",
    x = "Year", y = "Number of Cases", fill = "Disorder"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold")
  )
```
## Number of Mental Health Cases by Year
```{r}
ggplot(mental_not_death, aes(x = year, y = val, color = location_name)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::label_comma()) +
  labs(title = "Number of Mental Health Cases by Year",
       x = "Year", y = "Number of Cases", color = "Country") +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold"))
```


## 3.6 Event analysis

## Merging datasets
```{r}
merged_data <- merge(mental_not_death, gini_buckets, by.x = "location_name", by.y = "Country Name")

# Shows the distribution (median, spread, outliers) of depression rates for each inequality group.

# Basic boxplot
ggplot(merged_data, aes(x = gini_bucket, y = val)) +
  geom_boxplot(fill = "skyblue", alpha = 0.7) +
  labs(title = "Depression Rate by Gini Bucket",
       x = "Gini Coefficient Category",
       y = "Depression Rate (%)") +
  theme_minimal()


merged_data %>%
  filter(year.x %in% c(2010, 2015, 2020, 2023)) %>%  # Limit to a few years for clarity
  ggplot(aes(x = gini_bucket, y = val)) +
  geom_boxplot(fill = "lightgreen") +
  facet_wrap(~year.x) +
  labs(title = "Depression by Gini Bucket Over Time",
       x = "Gini Level",
       y = "Depression Rate") +
  theme_minimal()
#expenditure <- read_csv("health_spending.csv")

```

```{r}
  
# Step 1: Make sure WDI population data is ready
library(WDI)
library(dplyr)

pop_data <- WDI(country = "all", indicator = "SP.POP.TOTL", start = 2010, end = 2020) %>%
  rename(country = country, year = year, population = SP.POP.TOTL)

# Step 2: Join using location_name and year.x
merged_data <- merged_data %>%
  left_join(pop_data, by = c("location_name" = "country", "year.x" = "year"))

# Check population column
summary(merged_data$population)

# compute the rate:
merged_data <- merged_data %>%
  mutate(depression_rate = ifelse(!is.na(population), (val / population) * 100000, NA))

```
##  Spatial Analysis (Maps)
# ggplot2 + rnaturalearth for maps.
```{r}
#install.packages("sf")
#install.packages("rnaturalearth")
#install.packages("rnaturalearthdata")
library(rnaturalearth)
library(rnaturalearthdata)
library(sf)
```
```{r}
world <- ne_countries(scale = "medium", returnclass = "sf")

map_data <- merged_data %>% filter(year.x == 2015)

ggplot() +
  geom_sf(data = world, fill = "grey80", color = "white") +
  geom_sf(data = left_join(world, map_data, by = c("name" = "location_name")),
          aes(fill = depression_rate)) +
  scale_fill_viridis_c(option = "plasma") +
  theme_minimal() +
  labs(title = "Depression Rates by Country", fill = "Rate (%)")
``` 

```{r}

merged_data %>%
  filter(year.x %in% c(2010, 2015, 2013)) %>%
  ggplot(aes(x = gini_bucket, y = depression_rate)) +
  geom_boxplot(fill = "lightgreen") +
  facet_wrap(~year.x) +
  labs(title = "Depression by Gini Bucket Over Time",
       x = "Gini Level",
       y = "Depression Rate") +
  scale_y_continuous(labels = comma) +  # This disables scientific notation
  theme_minimal()
#expenditure <- read_csv("health_spending.csv")
```
```
```
Analyze the relationship between two variables.

```{r analysis}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

https://github.com/qgelena/Quantifying_a_social_problem/tree/main

## 5.2 Reference list

Use APA referencing throughout your document.
